A High Efficiency MAC Protocol for WLANs: Providing Fairness in Dense Scenarios

### 2015-11-09 responseLetter.tex

\documentclass[]{article}
\usepackage{amsmath}
\usepackage{mathtools}
\usepackage{fullpage}
\usepackage{graphicx}
\usepackage{color}
\usepackage[a4paper]{geometry}

\begin{document}

\title{Response letter}
  \author{Luis Sanabria-Russo, Jaume Barcelo, Boris Bellalta, Francesco Gringoli}

\date{\today}
\maketitle
\emph{Authors sincerely appreciate the time spent reading our work. Your insight propelled new considerations and discoveries that contributed to futher improve the paper. Thank you. \\\\This latter presents our responses to all the reviewers's comments. Additionnally, it provides an overview of the new contributions included in the paper. Authors kindly suggest to read it through, as some answers built on previous explanations. Again, thank you very much.\\}

In this new version of the paper, the more relevant additions are:
\begin{itemize}
	\item Completely redrawn set of figures with clearer explanations.
	\item A throughput comparison with similar reservation-like collision-free MAC protocols for IEEE 802.11 networks.
	\item Analysis of the impact of channel errors over CSMA/ECA$_{\text{Hys+FS}}$'s deterministic backoff.
	\item A new mechanism for reducing the length of collision-free schedules in CSMA/ECA$_{\text{Hys+FS}}$.
	\item The first experimental results of a real hardware implementation of CSMA/ECA with Hysteresis.
\end{itemize}

@@

	\subsection{Regarding ``Abstract"}
		Changes from \emph{``totally distributed"} to fully decentralised were made throughout the document.

	\subsection{Regarding ``Introduction"}
		\begin{itemize}
			\item {\bfseries What would be useful in the Introduction to specifically mention, is whether the proposed CSMA solution performs well when co-existing [perhaps with more greedy?] legacy CSMA/CA clients. The authors mention that throughput improves in co-existance scenarios, but it should be clear that it is not at the cost of better throughput for the nodes that are operating using the authors' proposed CSMA solution. This leaves less room for reader drifts.}
		\end{itemize}

		\begin{figure}[tb]
		\centering
			\includegraphics[width=0.45\linewidth,angle=-90]{figures/saturated/mixed/throughput-per-protocol/throughput-per-protocol-TON.eps}
			\caption{Average station throughput per MAC protocol in a saturated mixed network}
			\label{fig:coex}
		\end{figure}

		We appreciate the observation. CSMA/CA nodes are in fact very greedy. Nevertheless, in an mixed scenario example where half the nodes run CSMA/CA and half CSMA/ECA$_{\text{Hys+FS}}$, we observe a throughput enhancement for CSMA/CA when compared against a CSMA/CA-only network. This is now referred in the document in the following way:\\

@@

		`\emph{It is possible to see that for a low total number contenders ($N^{*}\leq 12$) CSMA/CA stations attain greater throughput than in a CSMA/CA-only network. Again, this is because in the mixed network setup the other $N^{*}/2$ contenders with CSMA/ECA$_{\text{Hys+FS}}$ use a deterministic backoff, leaving many empty slots between transmissions.}'

		`\emph{...for $N^{*}>12$ periods of collision-free operation and the aggregation performed with Fair Share allows CSMA/ECA$_{\text{Hys+FS}}$ to have larger channel time than CSMA/CA nodes, which throughput degrades even more than in CSMA/CA-only. These results suggests that a switch to CSMA/ECA$_{\text{Hys+FS}}$ can be beneficial for networks with high number of contenders.}'

		\begin{itemize}
			\item {\bfseries An implementation is crucial as well as the use of a measurement-based simulation environment. Good work. Mention in the introduction that the simulation is based on the COST simulation environment for greater credibility.}
		\end{itemize}

		We included a reference to a new customization of the COST simulator and incorporated a completely new hardware implementation Section.

@@

	\subsection{Regarding ``Related Work"}
		\begin{itemize}
			\item {\bfseries When stating the size $N$ of  collision slots, and that solutions that adopt such policies do not take care of situations where the number of nodes $M > N$, please give an example value for $N$. What number of M are we looking at? And is it reasonable to make the argument that $M > N$? If $N$ is large as it is, then the concern is not that $M > N$, but really can we service that many nodes? (at least not [reasonably] today) Perhaps a simple footnote could help clarify this.}
		\end{itemize}

		IEEE 8011.ax Task Group suggest simulation scenarios with $10's$ to $100's$ of stations for a single Access Point (AP). Based on the current minimum Contention Window (CW$_{\min}=16$) we can safely take an example of number of slots in a schedule, $N$ to be $16$ or $8$ slots. That is, collision-free schedules can be built with no more than $M\leq N$ contenders. When $M>N$ collisions will reappear, contributing to the observed overall throughput degradation.

		It is clear that a flexible or adjustable schedule is needed for collision-free operation with many contenders, as it is proposed in this work. Nevertheless, when $M$ is very big, say 200 contenders, and assuming $M\leq N$, the process of converging to a collision-free schedule will provoke some collisions, further degrading the overall throughput as the number of contenders increases. Besides, the increased time between successful transmissions will also degrade the average throughput of each node, but still keeping it higher than a CSMA/CA node.

@@

	\subsection{Regarding ``Section III"}
		\begin{itemize}
			\item {\bf It would be useful if the authors include a table with a list of important notations. Would help with readability.}
		\end{itemize}

		We now include a short table (Table I in the paper) with useful notation regarding backoff procedure variables and different CSMA/ECA configurations.

@@

		\begin{itemize}
			\item {\bfseries Section III.B, the aggregation process should be explained in more detail. Is the aggregation here similar to that used in 802.11n, where the level of aggregation depends on the transmission rate? How are packets aggregated? It is then mentioned in Section III.C that the packet is an MPDU and other details that indicate this protocol is applied for 802.11n+. Please clarify these details before then.}
		\end{itemize}

 		The authors apologise for raising any confusion. The aggregation performed with Fair Share does not depend on 		the rate. In fact, it only depends on the backoff stage at the moment of transmission. That is, stations at backoff 		stage $k$ will aggregate and transmit an A-MPDU frame with $2^{k}$ data frames. This is done with the sole 			purpose of providing fairness to CSMA/ECA$_{\text{Hys}}$ stations with different deterministic backoffs.

		We now first refer to Fair Share mentioning the following:\\

		`\emph{Fair Share is an A-MPDU aggregation mechanism that coupled with the collision-free schedule built by CSMA/ECA$_{\text{Hys}}$ is able to provide short-term fairness. However, it also improves the throughput since the aggregation process makes the packet transmission more efficient by reducing overheads. Furthermore, the level of aggregation provided by Fair Share depends on the buffer occupancy and the current backoff stage.}'

@@

		\begin{itemize}
			\item {\bfseries It is not clear in Section III.C what the authors mean by the "inferior" backoff stage. Please clarify or replace with a more descriptive word. Perhaps: subsequent, later, following, etc.}
		\end{itemize}

		We extended the explanation everywhere this could cause confusion. Authors kindly refer the Reviewer to Section III-C in the paper.

@@

	\subsection{Regarding ``Simulation"}
		\begin{itemize}
			\item {\bfseries The component lines in Figure 9 are not clear. They overlap. Perhaps in all graphs, align the legends with the order of their corresponding results. For example, the first legend item corresponds to the largest result (as in Figure 10). Also, this reader suggests to add a major line to separate the two parts of Figure 9 for more clarity --- this applies to other figures too. Also, more discussion on the values for the average backoff stage. What do large/small values mean? What kind of best pattern are we looking for? Results aren't as intuitive as other metrics.}
		\end{itemize}

		We completely regenerated the figures. All the legends follow the order of appearance of the curves in the Figure. We chose to group related figures together to save space and improve the structure of the paper.

		Backoff stages reflect the length of the collision-free schedule (if any) in CSMA/ECA$_{\text{Hys+FS}}$. When increasing the number of contenders in a network, it is expected to see an increase in the average backoff stage of each node given that a larger schedule length is needed to allocate those users (as can be observed in Figure 7a in the paper). Furthermore, in saturated scenarios with channel errors nodes will rapidly end up in the maximum backoff stage due to failed transmissions, and therefore use the largest deterministic backoff.

		On the other hand, in non-saturated traffic conditions we see an increase of the average backoff stage when the network approaches the saturation point (see Figure 8a in the paper). This is because CSMA/ECA and CSMA/CA reset their backoff stage to zero when the MAC queue empties.

		The ideal solution is to reach collision-free operation with the lowest backoff stage possible. This may affect the aggregation performed by Fair Share, but at the same decreases the time between successful transmissions and increases the channel utilization.

@@

		The use of the random backoff increases the number of collisions, which in turn increases the number of transmissions that reach the retransmission limit. As CSMA/ECA$_{\text{Hys+FS}}$ drops as many as $2^{k_{c}}$ packets (see Algorithm 3 in the paper), where $k_{c}$ is the backoff stage at the first transmission attempt, our proposal shows higher percentage of dropped packets than CSMA/CA for the aforementioned values of $N$.

		For $N>35$ nodes start to get saturated. This means that CSMA/ECA$_{\text{Hys+FS}}$ is able to provide collision-free operation for increasing periods of time. When $N=60$, the saturation point is reached and collisions are eliminated thanks to the construction of a collision-free schedule.

		\begin{itemize}
			\item {\bfseries This reader appreciates the level of thought and detail that was put into presenting these graphs. I think the authors could improve on the explanation of the results, as there are quite a bit of very interesting results and graphs presented. The explanation is quite good as is, but it often reads as a teaser, with a need for a little more detail to close the loop.}
		\end{itemize}

		Authors deeply appreciate the comment. We put extra effort in trying to enhance our explanations.

		We now provide a throughput comparison with the protocols presented in the related work section, consider the effects of channel errors, reviewed the algorithms presented, reworded explanations regarding non-saturated scenarios and coexistence (referring to Figure 8 and Figure 10).

		Hopefully we managed to transmit our ideas in a clearer way now.

@@

	\subsection{Regarding ``Other"}
		\begin{itemize}
			\item {\bf The authors did not discuss the impact of such things as hidden terminals on system performance (such as those discussed as early as [3]). Can the authors give insights on why that is the case? Perhaps a discussion outlining how this problem is out of scope, how it would possibly not impact performance results, how it is covered in other works... Some discussion to this avail. However, if space is a limitation, I would prefer the authors focus on deeper discussion of the results, rather than this discussion. Some mention or brief insight though would be appreciated.}
		\end{itemize}

		The influence of hidden terminals is briefly overviewed (see Section III-F). Nevertheless, a deep analysis of the impact of these type of terminals into our proposal is still lacking. This is expected to be assessed in a future work. Nevertheless, it is considered that hidden nodes will impact our proposal as they do in CSMA/CA networks.

		Hidden-terminals interrupt collision-free schedules as slot drifts or channel errors do. These will provoke an increase in the schedule length of CSMA/ECA$_{\text{Hys+FS}}$ contenders, leaving more free slots between its successful transmissions and therefore leveraging the effect of the hidden terminals. It is expected to see an increase in the backoff stage and throughput even at low number of contenders and hidden terminals. Much like the aggregated throughput under slot drifts figure (Figure 7e in the paper).

@@

\section{Reviewer 2}
Authors appreciate the Reviewer's comments regarding the scenarios proposed to test CSMA/ECA. Thanks to your insight we now provide a comprehensible set of results for different channel conditions.

	\subsection{Regarding ``Weaknesses"}
	\begin{itemize}
		\item {\bfseries The simulation results are obtained under simplistic scenarios where all nodes are within communication range and assuming perfect physical layer with no interference or channel errors.}
	\end{itemize}

	We appreciate your valuable comment. In fact, channel errors have a big impact over CSMA/ECA$_{\text{Hys+FS}}$. As failed transmissions due to channel errors are handled as collisions, CSMA/ECA$_{\text{Hys+FS}}$ nodes rapidly end up using the largest deterministic backoff. Even-though this means that more packets will be aggregated via Fair Share, it also implies an increase in the time between successful transmissions.

	We have incorporated channel errors to our results alongside a mechanism to leverage its impact over CSMA/ECA$_{\text{Hys+FS}}$'s deterministic backoff, namely Schedule Reset (SR).

	Schedule Reset looks at the empty slots between a node's consecutive transmissions and checks if it is possible to change the node's current deterministic backoff to occupy those empty slots. As CSMA/ECA$_{\text{Hys+FS}}$ uses a deterministic backoff $B_{\text{d}}=\lceil CW(k)/2\rceil -1$, where $CW(k)=2^{k}CW_{\min}$, the separation of empty slots should also be a power of two (please, refer to Section III-G in the paper for a complete overview of SR).

@@

	\begin{itemize}
		\item {\bfseries The proposed scheme is only compared against CSMA/CA.}
	\end{itemize}

	As our proposal is through to be an evolution of CSMA/CA, we compare all the performance metrics of CSMA/ECA$_{\text{Hys+FS}}$ against it. For a comparative analysis of collision-free protocols, authors respectfully refer the Reviewer to contributions like~\cite{L_MAC}.

	 In order to provide more context, average aggregated throughput and fairness curves for L-MAC and ZC-MAC are now included.

@@

	\begin{itemize}
		\item {\bfseries The authors do not propose a solution for the additional delays that may arise from Fair Share, which can be detrimental to time sensitive traffic.}
	\end{itemize}

	In order to decrease the time between successful transmission while using Fair Share, we propose Schedule Reset. Although it is able of reducing the time between successful transmissions by almost $43\%$, it is still higher than the same metric observed in CSMA/CA networks. Nevertheless, we now provide throughput and time between successful transmissions curves for CSMA/ECA$_{\text{Hys}}$. Although it is not completely fair, it outperforms CSMA/CA in all tested scenarios in terms of both throughput and delay.

@@

		Authors would like to thank the Reviewer for this observation.

		Even-though no real channel traces were used, our simulation of channel errors revealed that CSMA/ECA$_{\text{Hys+FS}}$ nodes get stuck with the largest deterministic backoff (also aggregating many packets with Fair Share). Nevertheless, for a very erroneous channel where, for instance $50\%$ of transmissions are corrupted by the channel, CSMA/ECA$_{\text{Hys+FS}}$'s minimum throughput will still be CSMA/CA's. Figure~\ref{fig:heavyErrors} shows the simulation results for throughput under the aforementioned conditions.

		\begin{figure}[tb]
		\centering
			\includegraphics[width=0.45\linewidth,angle=-90]{figures/tonFigs/heavyErrors.eps}
			\caption{Average throughput under heavy errors, $p_e=0.5$}
			\label{fig:heavyErrors}
		\end{figure}

		Authors plan to carefully develop the Reviewer’s suggestions in a new paper, where dense and spatially distributed topologies will be considered in detail. The study of the CSMA/ECA behavior in those topologies is not trivial, and we believe that the space that would require a detailed analysis of them makes it unfeasible to add it in this paper.

@@

	\begin{itemize}
		\item {\bfseries The proposed scheme is only compared against CSMA/CA. It would be beneficial to see how CSMA/ECA performs in comparison to other proposed zero collision schemes like Zero Collision MAC (ZC-MAC) and Learning MAC, which are referred to in the related work section of the paper.}
	\end{itemize}

		We now provide a throughput and fairness comparison with the protocols presented in the related work section.

%################################################################################

\section{Reviewer 3}
Authors want to acknowledge and thank you for the comments provided. Your insight on the effects of channel errors and/or node incorporation/withdrawal from the contention are very much appreciated. Our contribution now incorporates a mechanism for reducing the collision-free schedule length, which turned out to be a key feature for outperforming CSMA/CA under heavy channel errors (like in the new experimental results obtained from the real hardware implementation included in the paper).

@@

	\subsection{Regarding ``Weakness"}
		\begin{itemize}
			\item {\bfseries The related work section needs update. The discussion only includes two other studies, which is insufficient.}
		\end{itemize}

		The authors acknowledge the apparent reduced number of comparisons with other protocols. Nevertheless, was our consideration that attending CSMA/ECA-related issues were of higher priority due to space limitations. Nevertheless, authors enthusiastically suggest~\cite{L_MAC} as an appropriate survey on collision-free MAC protocols for WLANs. Furthermore, and as mentioned in the paper:\\

		`\emph{Performing time slot reservation for each transmission is a well known technique for increasing the throughput and mantanining Quality of Service (QoS) in TDMA schemes, like LTE~\cite{canoLTEcoexistence}. Applying the same concept to CSMA networks by modifying DCF's random backoff proceedure provides similar benefits~\cite{HE}. The following are MAC protocols for WLANs, decentralised and capable of attaining greater throughput than CSMA/CA by constructing collision-free schedules using reservation techniques. A survey of collision-free MAC protocols for WLANs is presented in~\cite{L_MAC}. In this paper we only overview those that are similar to CSMA/ECA.}'\\

		We attempt to increase the background on the subject by further constraining the category of our proposal. CSMA/ECA, as L-MAC and ZC-MAC are reservation-like protocols implemented for CSMA in WLANs. In the hopes of putting CSMA/ECA in context with L-MAC and ZC-MAC, we also included a throughput and fairness comparison.

@@

		\begin{itemize}
			\item {\bfseries The related work section discusses two MACs in a nice detailed way. However, a more comprehensive related work discussion on CSMA/CA is desired.}
		\end{itemize}

		Space limitations prevented the authors from including further details of CSMA/CA. Nevertheless, CSMA/ECA simply modifies the backoff mechanism of CSMA/CA, which is described by Algorithm 1 in the paper. Authors kindly suggest~\cite{bianchi2000performance} as a well-known study on CSMA/CA performance.

		\begin{itemize}
			\item {\bfseries So, when $2 < N < 7$, each node can send at most 1/8 of the slots. Shouldn't the aggregated throughput of N = 3 be around 1.5 times to the aggregate throughput of N = 2? The reviewer has similar question for later figures regarding to throughput.}
		\end{itemize}

		Thank you very much for raising this question, and give us the opportunity to further clarify Figure 1 with our response. To simplify the representation, in Figure 1 both the empty slots and those that contain a transmission have a similar duration. However, this is not the case, and empty slots are around 1000 times smaller than those slots that contain a transmission. Certainly, this is not shown in the figure, and to avoid any further confusion on this, we explain that in the caption.

### 2015-11-09 introProtocol.tex
Finished

CSMA/ECA~\cite{barcelo2008lba} is a fully distributed and collision-free MAC for WLANs. It differs from CSMA/CA in that it uses a deterministic backoff, $B_{\text{d}}=\lceil CW_{\min}/2\rceil-1$ after successful transmissions, where $CW_{\min}$ is the minimum contention window of typical value $CW_{\min}=16$. By doing so, contenders that successfully transmitted on schedule $n$, will transmit without colliding with other successful nodes in future cycles~\cite{HE}.

@@

Collision slots being orders of magnitude larger than empty slots degrade the network performance. When CSMA/ECA builds the collision-free schedule all contenders are able to successfully transmit more often, increasing the aggregated throughput beyond CSMA/CA's. Figure~\ref{fig:BECA}, \textcolor{red}{shows the average aggregated throughput of CSMA/ECA, CSMA/CA and two of the collision-free protocolos presented in Section~\ref{relatedWork}, namely L-ZC and L-MAC~\cite{L_MAC}}.

@@

Referring to Figure~\ref{fig:BECA}, CSMA/ECA is able to achieve an aggregated throughput that goes beyond CSMA/CA up until the number of contenders is greater than $B_{\text{d}}+1$ ($N=8$ in the case of the figure). Beyond this point, the network will have a mixed behaviour relating to backoff mechanisms: some nodes will successfully transmit and use a deterministic backoff while others will collide due to the lack of empty slots and return to a random backoff. As more contenders join the network, CSMA/ECA performance will approximate to CSMA/CA's. \textcolor{red}{This effect is also observed in L-ZC and L-MAC. It happens becasue the virtual cicle used by these protocolos ($M$ for L-ZC and $C$ for L-MAC in Section~\ref{relatedWork}) in lower than the number of contenders, $N$.}

@@

	Algorithm~\ref{alg:fullECA} describes CSMA/ECA$_{\text{Hys+FS}}$, Table~\ref{tab:glossary} provides a short list of notations used throughout the text, while an example of CSMA/ECA$_{\text{Hys+FS}}$ with four contenders is shown in  Figure~\ref{fig:ECA+Hyst}. In the figure the first outline indicates a collision between STA-3 and STA-4, which will provoke an increment on both station's backoff stage ($k\leftarrow k+1$). Once STA-4's random backoff expires, CSMA/ECA$_{\text{Hys+FS}}$ instructs the station to transmit $2^{k}$ packets, and then use a deterministic backoff, $B_{\text{d}}=\lceil CW(k)/2\rceil-1$. The same behaviour is followed by STA 3.

@@

	\textcolor{red}{Channel errors, hidden-nodes, or unsaturated traffic conditions will disrupt any collision-free schedule, generate collisions and push all stations's backoff stages to the maximum value very quickly (contributing to the delay increase). This issue is leveraged with the Schedule Reset mechanim, introduced in Section~\ref{schedReset}.}

@@

	\subsection{The effects of Aggregation}\label{effects-of-aggregation}
	Fair Share is an \textcolor{red}{A-MPDU} aggregation mechanism~\cite{A-MPDU} that coupled with the collision-free schedule built by CSMA/ECA$_{\text{Hys}}$ is able to provide short-term fairness. However, it also improves the throughput since the aggregation process makes the packet transmission more efficient by reducing overheads. Furthermore, the level of aggregation provided by Fair Share depends only on the buffer occupancy.

	The downside of Fair Share is that it may increase the time between two consecutive transmissions from the same node, which may affect negatively delay-sensitive applications such as gaming or high definition real-time video. \textcolor{red}{An example of the duration of a transmission, $T(l_{k})$, is provided by (\ref{eq:Ts}) in the following Section~\ref{ECA-bounds}}.

@@

	They correspond to the maximum and minimum achievable throughput without the possibility of collisions using Hysteresis and Fair Share. The ideal CSMA/ECA$_{\text{Hys+FS}}$ network uses the minimum schedule length that guarantees a collision-free operation. That is, with a schedule length of $C=2^{k}B_{\text{d}}+1$, where $k = \left\lceil log_{2}(N/(B_{\text{d}}+1))\right\rceil$. Using this minimum schedule length, $N$ nodes will be at the same backoff stage if $N\leq B_{\text{d}}+1$. Otherwise, $h = N-(C-N)$ nodes would occupy the $k$-th backoff stage and the other $N-h$ \textcolor{red}{nodes the $(k-1)$-th one}. The system throughput is computed as follows:

	\begin{equation}
			S =
				\begin{cases}\label{eq:bound}
					hs_{k}(l_{k}) + (N-h)s_{k-1}(l_{k-1}), & \text{if } N > B_{\text{d}} \\
					Ns_{k}(l_{k}), & \text{otherwise}
				\end{cases}
	\end{equation}

@@

	where $T_{\text{PHY}}=32~\mu$s is the duration of the PHY-layer preamble and headers, $T_{\text{sym}}=4~\mu$s is the duration of an OFDM (Orthogonal Frequency Division Multiplexing) symbol. SF is the \emph{service field} ($16$ bits), $\text{MD}$ is the \textit{MPDU Delimiter} ($32$ bits), MH is the \emph{MAC header} ($288$ bits), TB is the number of \emph{tail bits} ($6$ bits), $L_{\text{BA}}$ is the \emph{Block-ACK} length ($256$ bits) and $L_{\text{DBPS}}=256$ is the number of bits in each OFDM symbol. SIFS, DIFS and $\sigma_{e}$ values can be found in Table~\ref{tab:mac-params}.

@@

%	\textcolor{red}{\subsection{Real hardware implementation} 
%	Replacing the firmware of commercial WLAN cards with the Open Firmware for WiFi Networks, OpenFWWF~\cite{OpenFWWF}, allows the modification of time-sensitive operations in the MAC, like the backoff procedure. In Section~\ref{EDCA} we provide details regarding the tesbed, the firmware used and present results of real experiments using CSMA/ECA.}
%

	\textcolor{red}{\subsection{Channel errors}}\label{errorEffect}
	Failed transmissions due to channel errors are handled as collisions by CSMA/ECA$_{\text{Hys+FS}}$. Therefore, collision-free periods under this type of channel model are frequently interrupted. In order to accelerate the convergence to collision-free schedules in presence of channel errors CSMA/ECA$_{\text{Hys+FS}}$ instruct nodes to \emph{stick} to the current deterministic backoff even after \emph{stickiness} number of consecutive failed transmissions. Stickiness has been introduced to CSMA/ECA in~\cite{barcelo2011tcf}, where it is described and evaluated. Stickiness allows for a faster convergence towards a collision-free schedule, especially when performing under heavy channel errors.

@@

	\textcolor{red}{\subsection{Hidden terminals}}\label{hidden}
	One key characteristic of IEEE 802.11 devices is that their carrier sense range is at least two times greater than their data range~\cite{tuningCarrierSense}. In this situation, it could be expected that the impact of hidden nodes is low, as a given transmission can be only interfered by other transmissions from very distant nodes with energy levels not higher than the noise floor, which can be easily ignored. However, in specific deployments, where obstacles play also an important role on the propagation effects, hidden nodes may appear, causing asynchronous packet collisions~\cite{throughputUnderHiddenTerm}.

	In the case of CSMA/ECA$_{\text{Hys+FS}}$ WLANs, the same negative effects as in IEEE 802.11 WLANs are expected. In addition, collisions with hidden terminals will cause nodes to leave from any collision-free schedule, as channel errors do, thus continuously disrupting any attempt of collision-free operation.

@@

	\textcolor{red}{\subsection{Schedule Reset Mechanism}\label{schedReset}}
	CSMA/ECA$_{\text{Hys+FS}}$ instructs nodes to keep their current backoff stage after a successful transmission (resetting it to zero only if the node empties its MAC queue, see line~\ref{emptyQueue} in Algorithm~\ref{alg:fullECA}). This is done in order to increase the cycle length and provide a collision-free schedule for many contenders, which is desirable in dense scenarios.

@@

	%, coupled with an increase by one in the node's \emph{stickiness}~\cite{L_MAC2}. 
	%\footnote{Trailing waiting periods, like the SIFS, are composed of empty slots. These allow nodes to decrement their backoff counters even after a collision or successful slots. Without loss of generality, the stations in our examples and simulations decrement their respective backoff counters by one after these kind of slots (see example in Figures~\ref{fig:ECA} and~\ref{fig:ecaQoS}).}
		\begin{figure}[tb]
		\centering
			\includegraphics[width=\linewidth]{figures/tonFigs/scheduleReset.pdf}
			\caption{Example of the Schedule Reset mechanism. The lower row shows values of $t$ for a bitmap $b[t]$ of size $B_{\text{d}}=31$}
			\label{fig:scheduleReset1}
		\end{figure}

>>>>

CSMA/ECA~\cite{barcelo2008lba} is a fully decentralised and collision-free MAC for WLANs. It differs from CSMA/CA in that it uses a deterministic backoff, $B_{\text{d}}=\lceil CW_{\min}/2\rceil-1$ after successful transmissions, where $CW_{\min}$ is the minimum contention window of typical value $CW_{\min}=16$. By doing so, contenders that successfully transmitted on schedule $n$, will transmit without colliding with other successful nodes in future cycles~\cite{HE}.

@@

Collision slots being orders of magnitude larger than empty slots degrade the network performance. When CSMA/ECA builds the collision-free schedule all contenders are able to successfully transmit more often, increasing the aggregated throughput beyond CSMA/CA's. Figure~\ref{fig:BECA}, \textcolor{black}{shows the average aggregated throughput of CSMA/ECA, CSMA/CA and collision-free protocolos presented in Section~\ref{relatedWork}, namely ZC-MAC and L-MAC~\cite{L_MAC}}.

@@

Referring to Figure~\ref{fig:BECA}, CSMA/ECA is able to achieve an aggregated throughput that goes beyond CSMA/CA up until the number of contenders is greater than $B_{\text{d}}+1$ ($N=8$ in the case of the figure). Beyond this point, the network will have a mixed behaviour relating to backoff mechanisms: some nodes will successfully transmit and use a deterministic backoff while others will collide due to the lack of empty slots and return to a random backoff. As more contenders join the network, CSMA/ECA performance will approximate to CSMA/CA's. \textcolor{black}{This effect is also observed in ZC-MAC and L-MAC. It happens becasue the virtual cicle used by these protocolos ($M$ for ZC-MAC and $C$ for L-MAC in Section~\ref{relatedWork}) in lower than the number of contenders, $N$\footnote{L-MAC and ZC-MAC curves in Figure~\ref{fig:BECA} appear to yield the same aggregated throughput. This is because these protocols do not consider an increase in the backoff stage after a failed transmission, augmenting the collision probability beyond CSMA/CA's. An adaptation of the schedule length leverages this issue, which is preseted in~\cite{L_MAC} as \emph{Adaptive} L-MAC and L-ZC.}.}

@@

	Algorithm~\ref{alg:fullECA} describes CSMA/ECA$_{\text{Hys+FS}}$, Table~\ref{tab:glossary} provides a short list of notations used throughout the text, while an example of CSMA/ECA$_{\text{Hys+FS}}$ with four contenders is shown in  Figure~\ref{fig:ECA+Hyst}. In the figure the first outline indicates a collision between STA-3 and STA-4, which will provoke an increment on both stations' backoff stage ($k\leftarrow k+1$). Once STA-4's random backoff expires, CSMA/ECA$_{\text{Hys+FS}}$ instructs the station to transmit $2^{k}$ packets, and then use a deterministic backoff, $B_{\text{d}}=\lceil CW(k)/2\rceil-1$. The same procedure is followed by STA 3.

@@

	\textcolor{black}{Channel errors, hidden-nodes, or unsaturated traffic conditions will disrupt any collision-free schedule, generate collisions and push all stations' backoff stages to the maximum value very quickly (contributing to the delay increase). This issue is leveraged with the Schedule Reset mechanim, introduced in Section~\ref{schedReset}.}

@@

	\subsection{The effects of Aggregation}\label{effects-of-aggregation}
	Fair Share is an \textcolor{black}{A-MPDU} aggregation mechanism~\cite{A-MPDU} that coupled with the collision-free schedule built by CSMA/ECA$_{\text{Hys}}$ is able to provide short-term fairness. However, it also improves the throughput since the aggregation process makes the packet transmission more efficient by reducing overheads. Furthermore, the level of aggregation provided by Fair Share depends only on the buffer occupancy.

	The downside of Fair Share is that it may increase the time between two consecutive transmissions from the same node, which may affect negatively delay-sensitive applications such as gaming or high definition real-time video. \textcolor{black}{An example of the duration of a transmission, $T(l_{k})$, is provided by (\ref{eq:Ts}) in the following Section~\ref{ECA-bounds}}.

@@

	They correspond to the maximum and minimum achievable throughput without the possibility of collisions using Hysteresis and Fair Share. The ideal CSMA/ECA$_{\text{Hys+FS}}$ network uses the minimum schedule length that guarantees a collision-free operation. That is, with a schedule length of $C=2^{k}B_{\text{d}}+1$, where $k = \left\lceil log_{2}(N/(B_{\text{d}}+1))\right\rceil$. Using this minimum schedule length, $N$ nodes will be at the same backoff stage if $N\leq B_{\text{d}}+1$. Otherwise, $h = N-(C-N)$ nodes would occupy the $k$-th backoff stage and the other $N-h$ \textcolor{black}{nodes the $(k-1)$-th one}. The system throughput is computed as follows:

	\begin{equation}
			S =
				\begin{cases}\label{eq:bound}
					hs_{k}(l_{k}) + (N-h)s_{k-1}(l_{k-1}), & \text{if } N > B_{\text{d}}+1 \\
					Ns_{k}(l_{k}), & \text{otherwise}
				\end{cases}
	\end{equation}

@@

	where $T_{\text{PHY}}=32~\mu$s is the duration of the PHY-layer preamble and headers, $T_{\text{sym}}=4~\mu$s is the duration of an OFDM (Orthogonal Frequency Division Multiplexing) symbol. SF is the \emph{Service Field} ($16$ bits), $\text{MD}$ is the \textit{MPDU Delimiter} ($32$ bits), MH is the \emph{MAC header} ($288$ bits), TB is the number of \emph{Tail Bits} ($6$ bits), $L_{\text{BA}}$ is the \emph{Block-ACK} length ($256$ bits) and $L_{\text{DBPS}}=256$ is the number of bits in each OFDM symbol. SIFS, DIFS and $\sigma_{e}$ values can be found in Table~\ref{tab:mac-params}.

@@

%	\textcolor{black}{\subsection{Real hardware implementation} 
%	Replacing the firmware of commercial WLAN cards with the Open Firmware for WiFi Networks, OpenFWWF~\cite{OpenFWWF}, allows the modification of time-sensitive operations in the MAC, like the backoff procedure. In Section~\ref{EDCA} we provide details regarding the tesbed, the firmware used and present results of real experiments using CSMA/ECA.}
%

	\textcolor{black}{\subsection{Channel errors}}\label{errorEffect}
	Failed transmissions due to channel errors are handled as collisions by CSMA/ECA$_{\text{Hys+FS}}$. Therefore, collision-free periods under this type of channel model are frequently interrupted. In order to accelerate the convergence to collision-free schedules in presence of channel errors CSMA/ECA$_{\text{Hys+FS}}$ instruct nodes to \emph{stick} to the current deterministic backoff even after \emph{stickiness} number of consecutive failed transmissions. Stickiness has been introduced to CSMA/ECA in~\cite{barcelo2011tcf}, where it is described and evaluated. Stickiness allows for a faster convergence towards a collision-free schedule, especially when performing under heavy channel errors.

@@

	\textcolor{black}{\subsection{Hidden terminals}}\label{hidden}
	One key characteristic of IEEE 802.11 devices is that their carrier sense range is at least two times greater than their data range~\cite{tuningCarrierSense}. In this situation, the impact of hidden nodes can be considered to be low. This is because a given transmission could only be interfered by other transmissions from very distant nodes with energy levels not higher than the noise floor, which using modern radios can be easily solved by the capture effect. However, in specific deployments, where obstacles play also an important role on the propagation effects, hidden nodes may appear, causing asynchronous packet collisions~\cite{throughputUnderHiddenTerm}.

	In the case of CSMA/ECA$_{\text{Hys+FS}}$ WLANs, the same negative effects as in IEEE 802.11 WLANs are expected. In addition, collisions with hidden terminals will cause nodes to leave any collision-free schedule, as channel errors do, thus continuously disrupting any attempt of collision-free operation.

@@

	\textcolor{black}{\subsection{Schedule Reset Mechanism}\label{schedReset}}
	CSMA/ECA$_{\text{Hys+FS}}$ instructs nodes to keep their current backoff stage after a successful transmission (resetting it to zero only if the node empties its MAC queue, see line~\ref{emptyQueue} in Algorithm~\ref{alg:fullECA}). This is done in order to increase the cycle length and provide a collision-free schedule for many contenders, which is desirable in dense scenarios.

@@

	%, coupled with an increase by one in the node's \emph{stickiness}~\cite{L_MAC2}. 
	%\footnote{Trailing waiting periods, like the SIFS, are composed of empty slots. These allow nodes to decrement their backoff counters even after a collision or successful slots. Without loss of generality, the stations in our examples and simulations decrement their respective backoff counters by one after these kind of slots (see example in Figures~\ref{fig:ECA} and~\ref{fig:ecaQoS}).}
		\begin{figure}[tb]
		\centering
			\includegraphics[width=\linewidth]{figures/tonFigs/scheduleReset.pdf}
			\caption{Example of the Schedule Reset mechanism. The lower row shows values of $t$ for a bitmap $b[t]$ of size $|b[t]| = B_{\text{d}} + 1=32$}
			\label{fig:scheduleReset1}
		\end{figure}
